Figures to insert in text

### 2018-03-07 Chapters/Chapter02.tex
sdm review-text-after-read

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. Such model may be applied in all sort of applications because it would replicate human capacity to remember past experiences from clues of the present. For instance, when one is walking on a dark alley and is afraid of something, one cannot explain where ones fear come from. They just feel it. We may interpret this situation as clues of the present --- a dark alley --- recalling past experiences from memory and thus generating the scared feeling. Our memory is able to make a parallel between previous experiences and the clues. Although one has never been in the exactly same situation, ones brain makes an analogy and recognizes the danger. This flexibility into mapping one situation in another is an important human feature which is hard to replicate into computers.

It has been applied in many different fields, like pattern recognition \citep{norman2003modeling, rao1995natural}, noise reduction \citep{Meng2009}, handwriting recognition \citep{fan1997genetic}, robot automation \citep{Rajesh1998, mendes2008robot}, and so forth. \cite{Linhares2011} has showed that SDM respects the limits of short-term memory discussed by \citet{Miller1995} and \citet{Cowan2011}. Despite all those applications, there is not a reference implementation which would allow one to replicate the results published in a paper, to check the source code for details, and to improve it. Thus, even though intriguing results have been achieved using SDM, it requires great effort of researchers to improve someone's work.

It is our belief that such a tool could bring orders of magnitude more researchers and attention if they were able to use the model, at zero cost, with an easy to use high-level language such as python in an intuitive platform such as juypyter notebook. Neuroscientists interested in long-term memory storage should not have to worry about high-bandwidth vector parallel computation.  This new tool provides a ready to use system in which experiments can be executed almost as soon as they are designed and it may accelerate researches \citep{shen2014interactive}.

Our motivation was our own effort in order to run our models. As there is no reference implementation, we had to implement our own and run several simulations to ensure that our implementation was correct and bug free. Thus, we had to deviate from our main goal --- which was to test our models --- and to focus in the implementation itself. Furthermore, new members in our research group had to go through different source codes developed by former members.

Extensions of SDM has been used in many applications. For example, \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used a modified SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs. \citet{fan1997genetic} extended SDM with genetic algorithms. \citet{chada2016you} extended SDM creating the Rotational Sparse Distributed Memory (RSDM), which is used to modeling network motifs, dynamic flexibility, and hierarchical organization, all results from neuroscience literature.

The main contribution of this work is a reference implementation which yields (i) orders of magnitude gains in performance, (ii) has several backends and operations, (iii) has been validated against the mathematical model, (iv) is cross-platform, and (v) is easily extended to fulfill other research models. Our reference implementation may, hopefully, accelerate research into the model's dynamics and make it easier for readers to replicate any previous results and easily understand the source-code of the model.  Moreover, it is compatible with jupyter notebook and researchers may share their notebooks possibly accelerating the advances in their fields \citep{shen2014interactive}.

Other contributions have also been introduced, which include (i) a noise filtering approach, (ii) a supervised classification algorithm, (iii) and a reinforcement learning algorithm, all of them using only the original SDM proposed by Kanerva, i.e., with no additional mechanisms, algorithms, data structures, etc. Although some of our applications have already been explored in previous work \citep{Meng2009, fan1997genetic, rao1995natural}, all of them have done some adapting of SDM to their problems, and none of them have used just the ideas introduced by Kanerva. We have presented different approaches with no adaptations at all.

Finally, we have found an anomaly in one of Kanerva's prediction, which we believe is related to SDM capacity. We have also tested a generic reading operation proposed by professor Paulo Murilo (personal communication).

@@

\begin{tabular}{cp{\textwidth}}
  $n$ & Number of dimensions, i.e., $n=1,000$. \\
  $N$ & Size of the binary space --- $|\{0, 1\}^n| = 2^n$. \\
  $N'$ & Number of hard-locations samples from $\{0, 1\}^n$. Its typical value is 1,000,000, as suggested by \citet{Kanerva1988}. \\
  $H$ & Same as $N'$. \\
  $r$ & Access radius, i.e., when $n=1,000$ and $N'=1,000,000$, its typical value is $451$. This value is calculated to activate, on average, one thousand of $N'$. \\
  $\eta$ & A bitstring, usually a datum. \\
  $\eta_x$ & A clue $x$ bits away from $\eta$, i.e., $\text{dist}(\eta, \eta_x) = x$. \\
  $\xi$ & A bitstring, usually an address. \\
  $\text{dist}(x, y)$ & Hamming distance between $x$ and $y$ \\
  $\text{d}(x, y)$ & Same as $\text{dist}(x, y)$
\end{tabular}\\

@@

$\forall b \in \{ 0,1\} ^n$, we want an algorithm $A$ that yields the particular list of hard locations for $b$ and all hard locations respect the desired properties of the memory.  A reduction from measuring the distance to $10^6$ hard locations to a computation of $~ 10^3$ hard locations might yield astonishing performance gains, depending, of course, on our optimistic assumptions concerning existence and complexity of such algorithm.

At large scales of computing, the very ability to perform some experiments is a function of sheer performance. The horizon of experiments --- and possibly of knowledge --- expands \emph{as a function of computational demands}. More on this in my closing words.

>>>>

Sparse Distributed Memory (SDM) \citep{Kanerva1988} is a mathematical model of long-term memory that has a number of neuroscientific and psychologically plausible dynamics. This model may be applied in all sort of applications because of its incredible ability to closely reflect the human capacity to remember past experiences from clues of the present. For instance, when one is walking on a dark alley and is afraid of something, one cannot explain where one's fear come from. One just feels it. We may interpret this situation as clues of the present --- a dark alley; a giant metropolitan area; people going on about their lives mostly indifferent from each other; constrained routes ahead and behind you; etc. --- recalling past experiences from memory and thus generating the feeling. Our memory is able to make a parallel between previous experiences and the clues. Although one has never been in the exactly same situation, one's brain involuntarily makes an analogy and recognizes the possibility of danger. This flexibility into mapping one situation in another is an important human feature which is hard to replicate in computers.

SDM has been applied in many different fields, like pattern recognition \citep{norman2003modeling, rao1995natural}, noise reduction \citep{Meng2009}, handwriting recognition \citep{fan1997genetic}, robot automation \citep{Rajesh1998, mendes2008robot}, and so forth. \citet{Linhares2011} has showed that SDM respects the limits of short-term memory discussed by \citet{Miller1995} and \citet{Cowan2011}. Despite all those applications, there is not a reference implementation which would allow one to replicate the results published in a paper, to check the source code for details, and to improve it. Thus, even though intriguing results have been achieved using SDM, it requires great effort from researchers to build on top of previous work.

It is our belief that such a tool could bring orders of magnitude more researchers and attention if they were able to use the model, at zero cost, with an easy to use high-level language such as python, in an intuitive platform such as juypyter notebooks. Neuroscientists interested in long-term memory storage should not have to worry about high-bandwidth vector parallel computation.  This new tool provides a ready to use system in which experiments can be executed almost as soon as they are designed and it may accelerate research \citep{shen2014interactive}.

Our motivation was our own effort to run our models. As there is no reference implementation, we had to implement our own and run several simulations to ensure that our implementation was correct and bug free. Thus, we had to deviate from our main goal --- which was to test our hypothesis and explore the `idea space' --- and to focus in the implementation itself. Furthermore, new members in our research group had to go through different source codes developed by former members.

Extensions of SDM have been used in many applications. For example, \citet{Snaider2011} extended SDM to efficiently store sequences of vectors and trees.  \citet{Rajesh1998} used a modified SDM in an autonomous robot. \citet{Meng2009} modified SDM to clean patterns from noisy inputs. \citet{fan1997genetic} extended SDM with genetic algorithms. \citet{chada2016you} extended SDM creating the Rotational Sparse Distributed Memory (RSDM), which is used to modeling network motifs, dynamic flexibility, and hierarchical organization, all results from neuroscience literature.

The main contribution of this work is a reference implementation which yields (i) orders of magnitude gains in performance, (ii) has several backends and operations, (iii) has been validated against the mathematical model, (iv) is cross-platform, and (v) is easily extended to test new research ideas. Our reference implementation may, hopefully, accelerate research into the model's dynamics and make it easier for readers to replicate any previous results and easily understand the source-code of the model.  Moreover, it is compatible with jupyter notebook and researchers may share their notebooks possibly accelerating the advances in their fields \citep{shen2014interactive}.

Other contributions have also been introduced, which include (i) a noise filtering approach, (ii) a supervised classification algorithm, (iii) and a reinforcement learning algorithm, all of them using only the original SDM proposed by Kanerva, i.e., with no additional mechanisms, algorithms, data structures, etc. Although some of these applications have already been explored in previous work \citep{Meng2009, fan1997genetic, rao1995natural}, all of them have done some adapting of SDM to their problems, and none of them have used just the ideas introduced by Kanerva. We have presented different approaches with no adaptations whatsoever.

Finally, I have striven to provide a visual tour of the theory and application of SDM: whenever possible, detailed figures should tell the story --- or at least do the heavy lifting. In this study, we will see an anomaly in one of Kanerva's predictions, which I believe is related to SDM capacity. We will see tests of a generalized reading operation proposed by Physics Professor Paulo Murilo (personal communication).  We will see what happens when neurons --- and
all their information --- is simply and suddenly lost.  We will see whether information-theory can improve some of Kanerva's ideas.  From (basic) noise filtering to learning to play tic-tac-toe, we will review the entirety of Dr. Pentti Kanerva's proposal.

This time, however, it will be running on a computer.

@@

\begin{tabular}{cp{\textwidth}}
  $n$ & Number of dimensions, i.e., $n=1,000$. \\
  $N$ & Size of the binary space, $|\{0, 1\}^n| = 2^n$. \\
  $N'$ & Number of hard-locations samples from $\{0, 1\}^n$. Its typical value is 1,000,000, as suggested by \citet{Kanerva1988}. \\
  $H$ & Same as $N'$. \\
  $r$ & Access radius, i.e., when $n=1,000$ and $N'=1,000,000$, its typical value is $451$. This value is calculated to activate, on average, one thousandth of $N'$. \\
  $\eta$ & A bitstring, usually a datum. \\
  $\eta_x$ & A clue $x$ bits away from $\eta$, i.e., $\text{dist}(\eta, \eta_x) = x$. \\
  $\xi$ & A bitstring, usually an address. \\
  $\text{dist}(x, y)$ & Hamming distance between $x$ and $y$. \\
  $\text{d}(x, y)$ & Same as $\text{dist}(x, y)$.
\end{tabular}\\

@@

Consider the hypercube with $n$ dimensions.  We want to select a subset of its vertices with cardinality $2^{20}$ that is symmetrically distributed over the space. Afterwards, $\forall b \in \{ 0,1\} ^n$, we want an algorithm $A$ that yields the particular list of hard locations for $b$ and all hard locations respect the desired properties of the memory.

A reduction from measuring the distance to $2^{20}$ hard locations to a computation of $2^{10}$ hard locations might yield astonishing performance gains, depending, of course, on our optimistic assumptions concerning existence and complexity of such algorithm.  At large scales of computing, the very ability to perform some experiments is a function of sheer performance. The horizon of experiments --- and possibly of knowledge --- expands \emph{as a function of computational demands}. More on this in my closing words.

### 2018-03-07 Chapters/sdm2.tex
sdm review-text-after-read

Sparse Distributed Memory (SDM) is a mathematical model for cognitive memory published by \citet{Kanerva1988}. It introduces many interesting mathematical properties of $n$-dimensional binary space that, in a memory model, are psychologically plausible.  Most notable among these are the tip-of-the-tongue phenomenon, conformity to Miller's magic number \citep{Linhares2011} and robustness against loss of neurons.

The data and address space belong to binary space and are represented by a sequence of bits, called bitstrings. The distance between two bitstrings is calculated using the Hamming distance. It is defined for two bitstrings of equal length as the number of positions at which the bits are different. For example, $00110_{b}$ and $01100_{b}$ are bitstrings of length 5 and their Hamming distance is 2. One has to be careful when thinking intuitively about distance in SDM because the Hamming distance does not have the same properties of, say, the Euclidean distance in 3 dimensions, i.e., both follow triangle inequality ($d(A, B) \le d(A, C) + d(B, C)$), which Euclidean distance in 3 dimensions may be interpreted as ``if A is close to B, and B is close to C, then A is also close to C'' --- $d(A, B) \le r \text{and} d(B, C) \le r \Rightarrow d(A, C) \le 2r$ ---, but in SDM, although the inequality is also valid, two bitstrings would be close when, for instance, $r = 430$, so $2r = 860$ would cover all other bitstrings. Hence, it makes no sense to say that $A$ is also close to $C$. This difference in intuition may trick even experienced researchers when analyzing some situations.

@@

  \subfloat[$Q_{10}$]{\includegraphics[width=4.0in]{./images02/new-images/qn10.png}}
  \caption{Here we have $Q_n$, for $n \in$ \{3, 7, 10\}. Each node corresponds to a bitstring in $\{0,1\}^n$, and two nodes are linked iff the bitstrings differ by a single dimension.  A number of observations can be made here. First, the number of nodes grows as $2^n$ as $n$ grows; which makes the  space intractable as $n>>20$. Another interesting observation, better seen in the figures below, is that most of the space lies `at the center', at a distance of around 500 from any given vantage point.\label{hypercubes}}
\end{figure}

The space studied by Kanerva is also called the \emph{hypercube graph}, or $Q_n$, as in Figure \ref{hypercubes}. For a fixed $n \in \mathbb{Z}$, the graph $G = (V, E)$, in which $v \in V$ iff there is a bijective function $b: V\to \{0,1\}^n$ and $(v_i, v_j) \in E$ iff $H(b(v_i), b(v_j))=1$, where $H$ is the Hamming distance. That is, $n$-sized bitstrings correspond to nodes and edges exist between nodes $iff$ they flip a single bit.  Though Kanerva has derived many combinatorial properties of the space, additional results have been found in the graph-theoretical literature. A good survey is provided by \citet{harary1988survey}.

>>>>

Sparse Distributed Memory (SDM) is a mathematical model developed and suggested as a theory of human memory by Finish Scientist Pentti \citet{Kanerva1988}. It introduces many interesting mathematical properties of $n$-dimensional binary space that, in a memory model, seem to be remarkably psychologically plausible.  Most notable among these are the tip-of-the-tongue phenomenon, conformity to Miller's magic number \citep{Linhares2011} and robustness against loss of neurons.

The data --- and address space on which it is stored --- are represented by large sequences of bits, called \emph{bitstrings}. The \emph{Hamming distance} provides comparisons between bitstrings and is used as a metric for the system. The Hamming distance is defined for two bitstrings of equal length as the number of positions in which bits differ. For example, $00110_{b}$ and $01100_{b}$ are bitstrings of length 5 and their Hamming distance is 2.

@@

  \subfloat[$Q_{10}$]{\includegraphics[width=4.0in]{./images02/new-images/qn10.png}}
  \caption{Here we have $Q_n$, for $n \in$ \{3, 7, 10\}. Each node corresponds to a bitstring in $\{0,1\}^n$, and two nodes are linked iff the bitstrings differ by a single dimension.  A number of observations can be made here. First, the number of nodes grows as $O(2^n)$; which makes the space rapidly intractable. Another interesting observation, better seen in the figures below, is that most of the space lies `at the center', at a distance of around $n/2$ from any given vantage point.\label{hypercubes}}
\end{figure}

The space studied by Kanerva is also called the \emph{hypercube graph}, or $Q_n$, as in Figure \ref{hypercubes}. For a fixed $n \in \mathbb{Z}$, the graph $G = (V, E)$, in which $v \in V$ iff there is a bijective function $b: V\to \{0,1\}^n$ and $(v_i, v_j) \in E$ iff $H(b(v_i), b(v_j))=1$, where $H$ is the Hamming distance. That is, $n$-sized bitstrings correspond to nodes, and edges exist between nodes $iff$ they flip a single bit.  Though Kanerva has derived many combinatorial properties of the space, additional results have been found by the graph-theoretical community. A good survey is provided by \citet{harary1988survey}.

One has to be careful when thinking intuitively about distance in SDM because the Hamming distance does not have the same properties of, say, our 3-dimensional space.

Though both follow the triangle inequality ($d(A, B) \le d(A, C) + d(B, C)$), which in 3-d Euclidean distance may be loosely interpreted as ``if A is close to B, and B is close to C, then A is also close to C'' --- $d(A, B) \le r \text{ and } d(B, C) \le r \Rightarrow d(A, C) \le 2r$ ---, but in SDM, although the inequality is also valid, two bitstrings would be close when, for instance, $r = 430$, so $2r = 860$ would cover all other bitstrings. Hence, it makes no sense to say that $A$ is also close to $C$.

There are numerous, beautiful, counter-intuitive notions involved in this space. This difference in intuition may trick even experienced researchers when analyzing some situations.
